MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators

### 2024-02-11 Article.tex
rewrite intro; add outro; better connect MPI

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[1,\Letter]{Adam Šmelko}
\author[2]{Miroslav Kratochvíl}
\author[3,4,5,\Letter]{Vincent Noël}

\affil[1]{Department of Software Engineering, Charles University, Prague, Czech Republic}
\affil[2]{Luxembourg Centre for Systems Biomedicine, University of Luxembourg, Esch-sur-Alzette, Luxembourg}
\affil[3]{Institut Curie, Université PSL, F-75005, Paris, France}
\affil[4]{INSERM, U900, F-75005, Paris, France}
\affil[5]{Mines ParisTech, Université PSL, F-75005, Paris, France}
\maketitle

@@

\begin{corrauthor}
smelko@d3s.mff.cuni.cz, vincent.noel@curie.fr
\end{corrauthor}
\section*{Introduction}

MaBoSS \cite{stoll2012continuous, stoll2017maboss}\\
Why Boolean models are important\\
Why are they getting larger\\
Why are they need to be adapted for HPC environment\\

@@

In this section, we describe the most important aspects of MaBoSS algorithm in order to understand the optimizations described in the next section.

@@

Also, since the trajectory iteration is the main `hot loop' of the whole algorithm (it is executed $c \cdot u$ times), optimizing it as much as possible is imperative. The penalty for the processor handling a cache miss inside each loop can result in the orders-of-magnitude slowdown. This puts a big pressure on the choice of the data representation for the state $S$ and Boolean logic formulas. We discuss this topic in the next section.

@@

In the following section, we describe the implementation of MaBoSS.MPI and MaBoSS.GPU. We use this discussion to detail the specific optimization choices in the implementation of the algorithms.

\section*{Implementation}
\subsection*{MaBoSS.MPI}

\subsection*{MaBoSS.GPU}
Porting MaBoSS algorithm to GPUs poses an interesting task from the optimizations perspective. A simple rewrite to an annotation-based tool, such as OpenACC, may not be the most optimal strategy due to the usage of complex data structures, such as hash maps and expression trees, that are not well suited for GPU hardware. In this section, we describe a CUDA GPU implementation of the simulation step and the stats computation steps, and we detail the optimizations used to enable higher performance.

@@

@@

\subsection*{MaBoSS.MPI}
We developed an MPI parallel implementation of MaBoSS.

We simulated a cell cycle model by Sizek et al.\cite{sizek2019boolean} (Figure~\ref{fig:sizek_results}). A comparison of execution times of MaBoSS v2.5.3 and v2.4.0 using 40 cores and one million individual simulations showed a considerable (3×) speed-up between the pre- and post-optimisation versions, which is mostly due to optimised memory usage (figure 2, purple vs red). We used the same test with the MPI implementation of MaBoSS, using 20 cores for each MPI node, from 1 MPI node up to 192 MPI nodes. We can observe a very good speedup (>50x) up to 64 MPI nodes (1.28k cores), and a plateau for larger setups (Figure 7, green). This is probably due the very short simulation time (10s) for that setup and the minimum overhead for such a parallel simulation, and needs testing on larger simulations.

In order to explore larger model sizes, we created a synthetic model with 1000 nodes (Figure~\ref{fig:synthetic_results}). We simulated this model on 32 cores per MPI node, on 1 to 192 MPI nodes (32 to 6144 cores). With this dataset, the simulation time goes from 20 hours on 1 MPI node to 430 seconds on 192 nodes. As we hypothesised, with larger simulation times, a decline in the speedup was not observed as in the previous example 
The MaBoSS benchmarking measurements described in this section were performed on MareNostrum 4.

@@

In this section, we present the performance results of MaBoSS.GPU by comparing the runtimes of the CPU and GPU implementations on real-world and synthetic models. We also take a deeper look into the implementation by analyzing the overhead of the runtime compilation of the Boolean formulas.

For the benchmarks, we used $3$ real-world models with varying model sizes spanning from $10$ nodes to $133$. In order to test the scalability of the GPU implementation, we also created a synthetic models with up to $1000$ nodes. Synthetic models are designed in a way such that the length of each trajectory is predictable, and the models have no stable states. Also, the number of non-internal nodes is kept low (10 nodes) to enable the usage of the histogram optimization.

\subsubsection*{Measurements}

@@

@@

We did great work\\

\subsection*{Future works} 
bLA, BLA BLA

\begin{acknowledgements}
We thank Laurence Calzone and Gautier Stoll for their guidance and fruitful discussions.
\end{acknowledgements}

>>>>

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[1,\Letter]{Adam Šmelko}
\author[2]{Miroslav Kratochvíl}
\author[3,4,5]{Emmanuel Barillot}
\author[3,4,5]{Laurence Calzone}
\author[3,4,5,\Letter]{Vincent Noël}

\affil[1]{Department of Distributed and Dependable Systems, Charles University, Prague, Czech Republic}
\affil[2]{Luxembourg Centre for Systems Biomedicine, University of Luxembourg, Esch-sur-Alzette, Luxembourg}
\affil[3]{Institut Curie, Université PSL, F-75005, Paris, France}
\affil[4]{INSERM, U900, F-75005, Paris, France}
\affil[5]{Mines ParisTech, Université PSL, F-75005, Paris, France}
\maketitle

@@

\begin{corrauthor}
smelko@d3s.mff.cuni.cz, vincent.noel@curie.fr
\end{corrauthor}
\section*{Introduction}
Biological systems are large and complex, and understanding their behavior remains critical to designing new therapies for complex diseases such as cancer. A crucial approach in this endeavor is building computational models from existing knowledge and analyzing them to find intervention points and predict the efficacy of new treatments.
Many different frameworks have been used to describe biological systems, from quantitative systems of differential equations to more qualitative approaches such as Boolean models. While the former seems more adapted to represent complex behavior, such as non-linear dependencies, the latter is being increasingly used because it is capable of representing very large systems.
Many Boolean models have been built to describe biological systems to tackle a variety of problems: from understanding fundamental properties of cell cycle~\cite{faure2006cellcycle,sizek2019boolean} to more advanced properties of cancer~\cite{montagud2022prostate}.

Historically, the task of building Boolean models involved reading an extensive amount of literature and summarizing it in a list of essential components and their interactions. More recently, thanks to advances in databases listing such interactions~\cite{licata2020signor,turei2016omnipath} and experimental techniques providing information on a bigger number of components [rnaseq, single cell, …], the automatic methods have been designed to infer Boolean formulas from the constraints encoded in the knowledge and the experimental data~\cite{chevalier2020synthesis,benevs2023boolean} to build large Boolean models. While this effort faces many challenges, we believe it is the best way to study the large-scale complexity of biological systems. Furthermore, in order to analyze the dynamic properties of such large Boolean models, we need to develop efficient simulation methods. In this effort, adapting existing software to modern HPC computing architectures is an important goal.

Our work focuses on MaBoSS~\cite{stoll2012continuous, stoll2017maboss}, a stochastic Boolean simulator that performs estimations of state probability trajectories based on Markov chains. 
This paper discusses porting the MaBoSS simulation algorithm to two hardware architectures: GPU accelerators and large CPU clusters with an MPI parallel implementation.

@@

We begin with describing the most important aspects of MaBoSS algorithm in order to understand the optimizations described in the next section.

@@

Also, since the trajectory iteration is the main `hot loop' of the whole algorithm (it is executed $c \cdot u$ times), optimizing it as much as possible is imperative. The penalty for the processor handling a cache miss inside each loop can result in the orders-of-magnitude slowdown. This puts a big pressure on the choice of the data representation for the state $S$ and Boolean logic formulas. We discuss this topic in the Implementation section.

@@

Finally, we use this discussion to detail the specific optimization choices in the implementation of the algorithms.

\section*{Implementation}\label{sec:implementation}

In this section, we describe the high-level overview of MaBoSS.GPU and MaBoSS.MPI implementation.

\subsection*{MaBoSS.GPU}
Porting MaBoSS algorithm to GPUs poses an interesting task from the optimizations perspective. A simple rewrite to an annotation-based tool, such as OpenACC, may not be the most optimal strategy due to the usage of complex data structures, such as hash maps and expression trees, that are not well suited for GPU hardware. We divided the implementation into two parts: the bare \emph{simulation} and the trajectory \emph{statistics computation}.

@@

\subsection*{MaBoSS.MPI}
Since MaBoSS.MPI and MaBoSS.GPU projects were developed in parallel, the MPI implementation of MaBoSS is based on the original CPU code. Same as in GPU code, CPU implementation also assigns a single thread to a single trajectory. The MPI implementation simply extends this distribution of work among multiple MPI nodes, adding the second layer of parallelism.

In summary, MaBoSS.MPI evenly distributes individual trajectories first on MPI nodes and then on the corresponding CPU cores, assuming the MPI nodes are homogenous. Once the trajectories are simulated and the statistics are computed for each MPI node, the intermediate data are reduced into a final result using MPI collective operations.

% MaBoSS' CPU implementation already offered parallelism, allowing to use multiple CPU cores to compute the multiple trajectories. However, this implementation stays limited to a single machine CPU, and its scalability suffers from a very intensive memory usage. Furthermore, with the increasing number of nodes in Boolean models, the need for very large number of trajectories leads to large simulation time. 
% To tackle this, we decided to extend this implementation and add another layer of parallelism using MPI. In this implementation, the simulation of the trajectories is not only distributed on multiple cores, but on the multiple cores of multiple MPI nodes. By using this implementation, we are able to exploit the large CPU clusters available in HPC centers. 
% The implementation of this MPI implementation is straightforward, and consists in distributing the invidivual trajectories even further, first on MPI nodes and then on the CPU cores of the MPI node. Once the trajectories are simulated, the statistics are first computed for each MPI nodes, and then the global results is computer from the results of each MPI nodes.

@@

To evaluate the discussed optimizations, we present the results of performance benchmarks for MaBoSS.GPU and MaBoSS.MPI by comparing their runtimes against the original CPU implementation. To get the most accurate results, we used real-world models and synthetic models with varying sizes.

For the benchmarks, we used $3$ real-world models with varying model sizes spanning from $10$ nodes to $133$. In order to test the scalability of the GPU and MPI implementation, we also created a synthetic models with up to $1000$ nodes. Synthetic models are designed in a way such that the length of each trajectory is predictable, and the models have no stable states. Also, the number of non-internal nodes is kept low (10 nodes) to enable the usage of the histogram optimization.

The MPI implementation benchmarks were run on the MareNostrum 4 supercomputer.

@@

@@

\subsection*{MaBoSS.MPI}

Figure~\ref{fig:sizek_results} shows the efficiency of the MPI implementation on the Sizek model. We ran multiple suites, ranging from a single MPI node up to 192 nodes, each running 20 cores. We can observe a close-to-linear speedup of up to 64 MPI nodes (1.28k cores), and a plateau for larger suites (Figure 7, green). This can be explained by hitting a bottleneck in parallelization overhead and MPI communication cost when the problem is divided into too many small parts.

To stress the scalability of the implementation, we also used a synthetic model with 1000 nodes (see Figure~\ref{fig:synthetic_results}). We simulated this model on 32 cores per MPI node, on 1 to 192 nodes (32 to 6144 cores). With this configuration, the simulation time decreases from 20 hours on 1 MPI node to 430 seconds on 192 nodes. As we hypothesized, a decline in the speedup was not observed in larger simulations.

% We simulated a cell cycle model by Sizek et al.\cite{sizek2019boolean} (Figure~\ref{fig:sizek_results}). A comparison of execution times of MaBoSS v2.5.3 and v2.4.0 using 40 cores and one million individual simulations showed a considerable (3×) speed-up between the pre- and post-optimisation versions, which is mostly due to optimised memory usage (figure 2, purple vs red). We used the same test with the MPI implementation of MaBoSS, using 20 cores for each MPI node, from 1 MPI node up to 192 MPI nodes. We can observe a very good speedup (>50x) up to 64 MPI nodes (1.28k cores), and a plateau for larger setups (Figure 7, green). This is probably due the very short simulation time (10s) for that setup and the minimum overhead for such a parallel simulation, and needs testing on larger simulations.

% In order to explore larger model sizes, we created a synthetic model with 1000 nodes (Figure~\ref{fig:synthetic_results}). We simulated this model on 32 cores per MPI node, on 1 to 192 MPI nodes (32 to 6144 cores). With this dataset, the simulation time goes from 20 hours on 1 MPI node to 430 seconds on 192 nodes. As we hypothesised, with larger simulation times, a decline in the speedup was not observed as in the previous example 
% The MaBoSS benchmarking measurements described in this section were performed on MareNostrum 4.

% \begin{figure}%[tbhp]
% \centering
% \includegraphics[width=.8\linewidth]{Figures/large_model.pdf}
% \caption{Scalability results for synthetic model with 1000 nodes}
% \label{fig:synthetic_results}
% \end{figure}

@@

In this work, we presented two new implementations of MaBoSS, a continuous time Boolean model simulator, for HPC environments: the first implementation, MaBoSS.GPU, is designed to exploit the computational power of massively parallel GPU hardware, and the second implementation, MaBoSS.MPI, which enables MaBoSS to scale to many nodes of HPC clusters. We evaluated the performance of these implementations on real-world and synthetic models and demonstrated that both implementations are capable of providing significant speedups over the original CPU implementation. The GPU implementation shows $145-326\times$ speedup on real-world models, and the MPI implementation delivers a close-to-linear strong scaling on big models.

Overall, we believe the new MaBoSS extensions will enable the simulation of large, automatically generated models and will be a valuable tool for the systems biology community.

\subsection*{Future work}

As mentioned in the discussion section, the next steps in the development of MaBoSS.GPU would be to parallelize the iteration step of the simulation. This requires a different representation of Boolean formulas, which would not pose any thread divergence, probably by substituting runtime compilation with an in-memory data structure. This optimization could potentially decrease the register pressure created by holding the state data and thus increase the performance of huge models with hundreds of nodes.

Furthermore, many parallelization methods used in GPU implementation could be backported to the CPU implementation, increasing the overall usability of MaBoSS running on systems without GPU accelerators.

Finally, to enable the simulation of extra-large models with thousands of nodes, the two presented parallelization methods could be combined into a single high-performance solution by running GPU implementation on multiple MPI nodes.

\begin{acknowledgements}
We thank Gautier Stoll for his guidance and fruitful discussion.
\end{acknowledgements}
